# Exercise 2

```{r}
date()
```

The data set we are going to use for this exercise is found in the **data** folder named **learning2014.txt**. 

So let's read the file and move on to presenting it.

```{r}
lrn14 <- read.csv("data/learning2014.txt")
```

\

### 2.1 Structure and dimensions of the data:

The dataset presents results from a questionnaire where participants answered a set of questions regarding deep, surface and strategic learning. The dataset holds information
on participants gender, age, attitude and exam points.

The structure of the data is as follows:

```{r}
str(lrn14)
```
And the dimensions:

```{r}
dim(lrn14)
```
\

### 2.2 Graphical representation and summary of the data

The following graph shows a nice representation of all of the data, colour coded according to gender. We can see all of the variables plotted against each other as scatter plots, 
the independent variables distributions, and the correlations between variables. 

```{r}
library(GGally)
library(ggplot2)
p <- ggpairs(lrn14, mapping = aes(col=gender, alpha=0.3), lower = list(combo = wrap("facethist", bins = 20)))
p
```
\

The summary of all variables in the data can be viewed below.
```{r}
summary(lrn14)
```
\

### 2.3 Fitting a regression model 

For this part we will be fitting a regression model by using the exam points as our target variable, and the """!"#"#造!"造造#R!"造 as explanatory variables 
(mainly because I did not want to repeat what was done in datacamp). Below the model is created, and its summary printed out.

```{r}
model <- lm(points ~ gender + age + attitude, data = lrn14)
summary(model)
```
\
As we can see above in the summary, the gender and attitude are not significant in modelling the exam points (which is a good thing!), when compared to the attitude. So let's remove
gender and age from the model, essentially making the model a linear one:

```{r}
model <- lm(points ~ attitude, data = lrn14)
```
We can also easily plot this:

```{r}
qplot(attitude, points, data = lrn14) + geom_smooth(method = "lm")
```

\
### 2.4 Relationship between exam points and attitude

The model can be summarized as follows:

```{r}
summary(model)
```

As stated earlier, there was a correlation between attitudes and exam points, due to the low p-value. The residuals are quite high, meaning that we have quite spread out data, which
can also be seen from the model plot earlier. Due to the large spread, the multiple R-squared value is also quite low. But this does not mean necessarily that it is a bad model!

\
### 2.5 Model validation

The residuals vs fitted values, normal qq-plot and residuals vs leverage plots can be viewed below:

```{r}
par(mfrow=c(2,2))
plot(model, which=c(1,2,5))
```


The residuals vs leverage plot seems to be in such small scale on the x-axis, that everything seems fine. The same can be said of the residuals vs fitted plot, where the data seem quite 
evenly clumped together. However, from the qq-plot we can see that the data deviates somewhat in the beginning and end from the line. This means that the normality assumption questionable. 
